The curious world of constructive mathematics

### 2022-09-04 algar-lecture-notes.tex
BHK, computability and geometric interpretation

\title{The curious world of constructive mathematics}
\author{Ingo Blechschmidt}
%\email{iblech@speicherleck.de}

@@

@@

Relinquishing the axiom of choice and the law of excluded middle also
allows us to explore axioms and notions which are incompatible with
these classical laws, such as mathematical settings in which all
functions are continuous or in which the intuitive idea of a ``generic
ring'' can be put on a firm basis.
\end{abstract}

{
\renewcommand{\newpage}{\ }
\renewcommand{\vfill}{\ }
\renewcommand{\vfil}{\ }
{\vspace*{-6em}\centering\includegraphics[width=0.3\textwidth]{phantoms}\par}
\thispagestyle{empty}\enlargethispage{7em}
\maketitle
}

\setcounter{tocdepth}{1}
{
\renewcommand{\newpage}{\ }
\tableofcontents
}

@@

\begin{intro}
\it
This lecture provides a first glimpse of constructive mathematics with a
focus on applications of constructive mathematics and on providing
intuition for quickly discerning which techniques and results hold
constructively.
\end{intro}

@@

Of the many axioms and inference rules of classical logic, exactly one is
responsible for enabling unconstructive arguments, namely the \emph{principle
of excluded middle}:
\[ \varphi \ \vee\ \neg\varphi. \]
The first proof above used this principle in its very first step. In
constructive mathematics, we abstain from this principle; we build
constructive mathematics on \emph{intuitionistic logic}, which contains neither
this principle nor the (equivalent) \emph{principle of double negation
elimination} stating~$\neg\neg\varphi \Rightarrow \varphi$, and insofar as we
layer a set theory on top of our logical foundation, we abstain from the axiom
of choice (which in presence of other common set-theoretical axioms implies the
principle of excluded middle, see Exercise~\ref{ex:diaconescu}). As a
consequence, we cannot generally reason by contradiction in constructive mathematics.

In constructive mathematics, we do \emph{not} claim that the principle of
excluded middle is false. \marginpar{\dbend}
Indeed, intuitionistic logic is downwardly compatible
with classical logic (every intuitionistic proof is a fortiori also a classical
proof), and some special instances of the principle of excluded middle are
intuitionistically verifiable (an example is given in
Proposition~\ref{prop:discreteness}). Instead, in constructive mathematics we
merely do not use the principle of excluded middle.

@@

A rumor about constructive mathematics states that constructively, the term
``contradiction'' would be generally forbidden. This is false. In fact, we need
to distinguish between two distinct figures of proof which are often conflated
in classical informal mathematics:

@@

Arguments of the first form are proper proofs by contradiction and hence not
generally acceptable in constructive mathematics. What is established is
only~$\neg(\neg\varphi)$, the impossibility of~$\neg\varphi$; constructively,
this is weaker than a positive affirmation of~$\varphi$. Though there are
situations in which~$\neg\neg\varphi$ implies~$\varphi$, generally it does not
and hence intuitionistically arguments of the first form fail to
estabilish~$\varphi$.

@@

\subsection*{The Brouwer--Heyting--Kolmogorov interpretation}

The notion of witnesses is compositional in nature and starts out with witnesses for \emph{atomic
statements}, those not built from substatements using the logical
connectives~$\wedge,\vee,\Rightarrow$ or the quantifiers~$\forall,\exists$. For
instance, in formal number theory, the atomic statements are of the form
``$s = t$'',
where~$s$ and~$t$ are terms for natural numbers; statements of this form are so
simple that we can check them directly so that witnesses don't need to supply
particular additional information. We hence decree that true atomic statements
have trivial witnesses and that false atomic statements have no witnesses at
all.

Table~\ref{table:bhk} explains what witnesses of compound statements are. For
instance, a witness of a statement of the form
\[ \forall n \? \NN\_ \bigl(\varphi(n) \Rightarrow \psi(n)\bigr) \]
is a rule explaining how, for every natural number~$n : \NN$, witnesses
for~$\varphi(n)$ give rise to witnesses for~$\psi(n)$.

\begin{table}
  \centering
  \small
  \renewcommand{\arraystretch}{1.3}
  \begin{tabular}{@{}rp{4.3cm}p{5.3cm}@{}}
    \toprule
    & {classical logic} & {intuitionistic logic}
    \\\midrule
    statement $\varphi$ & The statement $\varphi$ holds. & We have a witness for $\varphi$. \\
    $\bot$ & A contradiction holds. & We have a witness for a contradiction. \\
    $\varphi \wedge \psi$ & $\varphi$ and $\psi$ hold. & We have a witness for~$\varphi$ and for~$\psi$. \\
    $\varphi \vee \psi$ & $\varphi$ or $\psi$ hold. & We have a witness for~$\varphi$ or for~$\psi$. \\
    $\varphi \Rightarrow \psi$ & If~$\varphi$ holds, then so does~$\psi$. &
    We can (uniformly) construct witnesses for~$\psi$ from witnesses
    for~$\varphi$. \\
    $\neg\varphi$ &
      $\varphi$ does not hold. &
      There is no witness for~$\varphi$. \\
    $\forall x\?X\_ \varphi(x)$ & For all $x \? X$ it holds that~$\varphi(x).$ &
      We can (uniformly) construct, for all~$x \? X$, witnesses for~$\varphi(x)$. \\
    $\exists x\?X\_ \varphi(x)$ & \raggedright There is at least one~$x \? X$
    such that~$\varphi(x)$ holds. & {\raggedright
      We have a~$x \? X$ together with a witness for~$\varphi(x)$.} \\
    \bottomrule
  \end{tabular}
  \caption{\label{bhk}Informale rekursive Definition des Belegbegriffs.}
\end{table}

\begin{ex}
According to the Brouwer--Heyting--Kolmogorov interpretation, the principle of
excluded middle states that there is a rule for producing, for every~$\varphi$,
either a witness for~$\varphi$ or a witness for~$\neg\varphi$. The BHK
interpretation does not specify which kind of rule ...
Unter dieser Interpretation meint das Prinzip vom ausgeschlossenen Dritten, dass wir für jede
Aussage Beleg für sie oder ihre Negation haben. Das ist aber offensichtlich
nicht der Fall.
\end{ex}

%\begin{bsp}
Die Interpretation von~$\neg\neg\varphi$ ist, dass es keinen Beleg
für~$\neg\varphi$ gibt. Daraus folgt natürlich noch nicht, dass wir tatsächlich
Beleg für~$\varphi$ haben; gewissermaßen ist eine solche Aussage~$\varphi$ nur
"`potenziell wahr"'.
%\end{bsp}

%\begin{bsp}
Wenn wir wissen, dass sich unser Haustürschlüssel irgendwo in der
Wohnung befinden muss (da wir ihn letzte Nacht verwendet haben, um die Tür
aufzusperren), wir ihn momentan aber nicht finden, so können wir konstruktiv
nur folgende doppelt negierte Aussage vertreten:
\[ \neg\neg (\exists x\_ \text{der Schlüssel befindet sich an Position~$x$})
\]
%\end{bsp}

%\begin{bsp}
Wir stehen im Supermarkt und erinnern uns, dass wir unbedingt
gewisse Zutaten einkaufen müssen. Leider fällt uns momentan keine einzige der
Zutaten mehr ein. Dann können wir zwar die Aussage, dass die Menge der zu
besorgenden Zutaten nicht leer ist, vertreten, nicht jedoch die stärkere
Aussage, dass diese Menge ein Element enthält.
%\end{bsp}

%\begin{bsp}[\cite{sigfpe:katemoss,bbc:katemoss}]
Es war ein Video aufgetaucht, dass Kate Moss beim Konsumieren von Drogen zeigte,
und zwar entweder solche von einem Typ~A oder solche von einem Typ~B. Welcher
Typ aber tatsächlich vorlag, konnte nicht entschieden werden. Also gab es für
keine der beiden Straftaten einen Beleg, Kate Moss wurde daher nicht
strafrechtlich verfolgt.
%\end{bsp}

%\begin{bem}Die Formulierung mit dem generischen \emph{Wir} in Tafel~\ref{bhk}
ist etwas irreführend. Wie auch in klassischer Mathematik hängen
intuitionistische Urteile nicht von \emph{uns} oder anderen Mathematikern ab.
Aussagen, die bisher noch nicht konstruktiv bewiesen wurden, können
durchaus einen (noch unbekannten) Beleg besitzen. Eine genauere Diskussion
findet sich etwa in~\cite[Seite~42f.]{kohlenbach:applprooftheory}. Die
Brouwer--Heyting--Kolmogorov-Interpretation kann im Rahmen der
\emph{Realisierbarkeitstheorie} formalisiert werden~\cite{bauer:realizability};
ohne eine solche Formalisierung gibt es gewisse weitere
Probleme~\cite{artemov:bhk,sanz-piecha:critical-bhk,dalen:bhk}.
%\end{bem}

\subsubsection*{\ldots{} über Berechenbarkeit}

Es gibt noch eine zweite Interpretation, die beim Verständnis der konstruktiven
Sichtweise sehr hilfreich ist:
%\listedmotto{konstruktiv-gdw-computer}{Eine Aussage gilt genau dann konstruktiv, wenn es ein
%Computerprogramm gibt, welches sie in endlicher Zeit bezeugt.}
Etwa ist mit dieser Interpretation klar, dass die formale Aussage
\[ \forall n \in \NN\_ \exists p \geq n\_ \text{$p$ ist eine Primzahl}, \]
eine Formulierung der Unendlichkeit der Primzahlen, auch konstruktiv
stimmt: Denn man kann leicht ein Computerprogramm angeben, das eine natürliche
Zahl~$n$ als Eingabe erwartet und dann, etwa über die Sieb-Methode von
Eratosthenes, eine Primzahl~$p \geq n$ produziert (zusammen mit einem Nachweis,
dass~$p$ tatsächlich prim ist).

Das Motto kann man tatsächlich zu einem formalen Theorem
präzisieren, das ist Gegenstand der gefeierten
Curry--Howard-Korrespondenz (Abschnitt~\ref{curry-howard}).

@@

{\noindent\scriptsize\emph{Remark.} The set~$\NN_\infty$ is the \emph{one-point
compactification} of~$\NN$. A sensible definition of it in constructive mathematics
is as the set of decreasing binary sequences~$(x_0,x_1,x_2,\ldots)$. The
naturals embed into~$\NN_\infty$ by mapping~$n$ to the sequence~$1^n 0^\omega =
(1,\ldots,1,0,\ldots,0)$, and an element not in the image of this embedding
is~$\infty \defeq 1^\omega = (1,1,\ldots)$. Assuming the law of excluded
middle (or already weaker principles), every element of~$\NN_\infty$ is of one
of these two forms. Martín Escardó has worked extensively on unexpected
instances of the principle of omniscience for searchable sets
like~$\NN_\infty$~\cite{escardo:omniscience1,escardo:omniscience2,escardo:omniscience}.\par}
\end{enumerate}
\end{exercise}

@@

\begin{exercise}[Brouwerian counterexamples]
Show that each of the following statements implies the law of excluded middle,
hence is not available in constructive mathematics.
\begin{enumerate}
\item Every ideal of~$\ZZ$ is finitely generated.

@@

{\noindent\scriptsize\emph{Note.} A set~$X$ is \emph{Kuratowski-finite} if and
only if, for some number~$n \in \NN$, there is a surjective map~$[n] \to X$,
where~$[n] = \{ 0,1,\ldots,n-1 \}$. More briefly, a set~$X$ is
Kuratowski-finite iff its elements can be enumerated: $X = \{x_1,\ldots,x_n \}$.\par}
\end{enumerate}
Show that the following statement implies Markov's principle (from
Exercise~\ref{ex:tautologies}):
\begin{enumerate}
\addtocounter{enumi}{3}
\item Every real number which is not zero is invertible.
\end{enumerate}
\noindent
More Brouwerian counterexamples can be
found in~\cite{mandelkern:brouwerian-counterexamples} and
in~\cite{mines-richman-ruitenburg:constructive-algebra}.
\end{exercise}

\begin{exercise}[Diaconescu's theorem]
\label{ex:diaconescu}
The axiom of choice can be put as: ``Every surjective map has a
section.'' (A \emph{section}~$s$ to a surjective map~$f$ is a map in the other direction such that~$f \circ s =
\mathrm{id}$.) A theorem of Diaconescu states that
the axiom of choice implies the law of excluded middle. To this end, let~$\varphi$ be a
statement and consider the subsets
\begin{align*}
  U &= \{ x \in X \,|\, (x = 0) \vee \varphi \} \\
  V &= \{ x \in X \,|\, (x = 1) \vee \varphi \}
\end{align*}
of the discrete set~$X \defeq \{ 0,1 \}$.
\begin{enumerate}
\item Verify that~$U = V$ if and only if~$\varphi$.
\item Using that~$x = y \vee x \neq y$ for all elements~$x,y \in X$, show that
the existence of a section of the surjective map
\[ \begin{array}{@{}rcl@{}}
  X &\longrightarrow& \{U,V\} \\
  0 &\longmapsto& U \\
  1 &\longmapsto& V
\end{array} \]
implies~$\varphi \vee \neg\varphi$.
\end{enumerate}
\end{exercise}

\vfill
{\small
\begin{verbatim}
- What constructive mathematics is about: finer distinctions, algorithmic
  interpretation, geometric interpretation (including teaser on why we
  might care)
- Proof by and proof of a contradiction: powers of irrational numbers,
  discreteness of N, Q, Qbar, failure of discreteness of R
- Positive information from negative? de Morgan, minimums of sets of
  natural numbers
- Relation to classical mathematics
- Finite sets and the axiom of choice
- Axiomatic freedom: all functions continuous, all functions
  computable, generic prime ideal
\end{verbatim}
}

\chapter{On the constructive content of classical proofs}

\begin{intro}
Decades of experience in constructive mathematics show: \emph{Most results
in classical mathematics, even those whose proof rests on
non-constructive principles like the axiom of choice or the law of
excluded middle, have a hidden constructive core.} With a mix of
experience, seasoned tools and general metatheorems, this constructive
content can be extracted from classical proofs. In this way we obtain
constructive reformulations of classical results, especially
if they are of a sufficiently concrete nature.

@@

What is the meaning of these non-effective proofs? Theorem~\ref{thm:dickson}
claims the existence of a finite object with a decidable property (a
pair~$(i,j)$ such that~$f(i) \leq f(j)$), but the given proof employs
transfinite methods and gives no indication how we could compute or otherwise
find this object. Instead, the classical proof asks us to grasp the infinitude
of all values of~$f$ so we can determine the minimum. The issue is even more
pronounced with Theorem~\ref{thm:surjective-matrix}, since the existence of
maximal ideals in nontrivial rings requires the axiom of
choice~\cite{scott:prime-ideals,hodges:krull,banaschewski:krull,erne:krull,howard-rubin:ac}.
To add one more conundrum: Assume that we have a classical proof, using the law
of excluded middle and the axiom of choice, that some given Turing machine
terminates. Can we then constructively accept that the machine will halt? Do we
have an upper bound for the number of computational steps the machine carries
out before halting?

@@

\item \emph{Eliminating the law of excluded middle by the double-negation
translation and its variants.} At the price of
slightly modifying the asserted claim, the law of excluded middle
can always be mechanically eliminated from a given proof. This
elimination procedure is facilitated by the \emph{double-negation translation}
reviewed below. In some cases, a refined translation even allows to preserve the asserted
claim exactly; this technique is variously known as \emph{Friedman's trick},
\emph{nontrivial exit continuation} or (the baby version of) \emph{Barr's
theorem}. To cite a concrete instance of this phenomenon, classical~\textsc{zf}
set theory is conservative over its intuitionistic cousin~\textsc{izf}
for~$\Pi^0_2$-statements (statements of the form~$\forall\ldots\forall\_
\exists\ldots\exists\_ \%$, where all quantifiers in the final~``$\%$''
are bounded).

@@

Noticeably missing in this list is any technique for eliminating uses of the
\emph{powerset axiom} stating that the collection of all subsets of a given
set is again a set. While this axiom is uncontested by ordinary constructive
mathematics and doesn't receive nearly as much philosophical attention as the
axiom of choice or the law of excluded middle, it is this axiom which actually
and substantially increases logical strength. While~\textsc{zfc}, \textsc{zf}
and \textsc{izf} are equiconsistent (and in fact verify the same
arithmetical~$\Pi^0_2$-statements), systems without it such as Kripke--Platek
set theory~(\textsc{kp}) or constructive Zermelo--Fraenkel set
theory~(\textsc{czf}) are much weaker. We will not discuss this curious state
of affairs; references include~\cite{crosilla:predicativity,aczel-rathjen:cst}.

\section{The double-negation translation}

\section{Barr's theorem}

@@

\section{Examples from quadratic form theory}

@@

\chapter{Generic models and their applications}

@@

\section{A fantastical ring}
\newpage

\section{The generic prime filter}
\newpage

\section{Explicit construction}
\newpage

>>>>

\title{The curious world of constructive mathematics}
\author{Ingo Blechschmidt \vspace*{-2em}}
%\email{iblech@speicherleck.de}

@@

\urlstyle{sf}

@@

Relinquishing the axiom of choice and the principle of excluded middle also
allows us to explore axioms and notions which are incompatible with
these classical laws, such as mathematical settings in which all
functions are continuous or in which the intuitive idea of a ``generic
ring'' can be put on a firm basis.
\end{abstract}

{
\renewcommand{\newpage}{\ }
\renewcommand{\vfill}{\ }
\renewcommand{\vfil}{\ }
\enlargethispage{7em}
{\vspace*{-6em}\centering\includegraphics[width=0.3\textwidth]{phantoms}\par}
\maketitle
}

\setcounter{tocdepth}{1}
{
\renewcommand{\newpage}{\ }
\tableofcontents
}
\thispagestyle{empty}

@@

\begin{intro}
\it
This lecture provides a first glimpse of constructive mathematics with a
focus on applications of constructive mathematics and on providing
intuition for quickly discerning which techniques and results hold
constructively.

This account is an update and translation of~\cite[Section~1]{blechschmidt:pizzaseminar}
and references
include~\cite{bauer:five-stages-video,bauer:five-stages-article,avigad:constructive,sep:intuitionistic-logic,dummett:basis,melikhov:intuitionistic-logic,streicher:constructive}
or more
specifically~\cite{mines-richman-ruitenburg:constructive-algebra,lombardi-quitte:constructive-algebra}
for constructive algebra and~\cite{bishop-bridges:bible} for constructive
analysis.
\end{intro}

@@

Of the many axioms and inference rules of classical logic, exactly one is
responsible for enabling unconstructive arguments, namely the \emph{principle
of excluded middle}:
\[ \varphi \ \vee\ \neg\varphi. \]
The first proof above used this principle in its very first step. In
constructive mathematics, we abstain from this principle; we build
constructive mathematics on \emph{intuitionistic logic}, which contains neither
this principle nor the (equivalent) \emph{principle of double negation
elimination} stating~$\neg\neg\varphi \Rightarrow \varphi$, and insofar as we
layer a set theory on top of our logical foundation, we abstain from the axiom
of choice (which in presence of other common set-theoretical axioms implies the
principle of excluded middle, see Exercise~\ref{ex:diaconescu}) and from Zorn's lemma. As a
consequence, we cannot generally reason by contradiction in constructive
mathematics, and to demonstrate the existence of an object
it is not enough that its nonexistence would entail a contradiction.

Importantly, in constructive mathematicics we do \emph{not} claim that the principle of
excluded middle is false. \marginpar{\dbend}
Indeed, intuitionistic logic is downwardly compatible
with classical logic (every intuitionistic proof is a fortiori also a classical
proof), and some special instances of the principle of excluded middle and with
it some special instances of proof by contradiction are
intuitionistically verifiable (an example is given in
Proposition~\ref{prop:discreteness}). Instead, in constructive mathematics we
merely do not use the principle of excluded middle.

@@

A rumor about constructive mathematics states that constructively, the term
``contradiction'' would be generally forbidden. This rumor is false. In fact, we need
to distinguish between two distinct figures of proof which are often conflated
in classical informal mathematics:

@@

Arguments of the first form are proper proofs by contradiction and hence not
generally accepted in constructive mathematics. What they establish is
only~$\neg(\neg\varphi)$, the impossibility of~$\neg\varphi$; constructively,
this is weaker than a positive affirmation of~$\varphi$. Though there are
situations in which~$\neg\neg\varphi$ implies~$\varphi$, generally it does not
and hence intuitionistically arguments of the first form fail to
estabilish~$\varphi$.

@@

\subsection{The Brouwer--Heyting--Kolmogorov interpretation}

The notion of witnesses is compositional in nature and starts out with the notion of witnesses for \emph{atomic
statements}, those not built from substatements using the logical
connectives~$\wedge,\vee,\Rightarrow$ or the quantifiers~$\forall,\exists$. For
instance, in formal number theory, the atomic statements are of the form
``$s = t$'',
where~$s$ and~$t$ are terms for natural numbers; statements of this form are so
simple that we can check them directly so that witnesses don't need to supply
particular additional information. We hence decree that true atomic statements
have trivial witnesses and that false atomic statements have no witnesses at
all.

Table~\ref{table:bhk} explains how compound statements should be witnessed. For
instance, a witness of a statement of the form
\[ \forall n \? \NN\_ \bigl(\varphi(n) \Rightarrow \psi(n)\bigr) \]
is a rule explaining how, for every natural number~$n : \NN$, witnesses
for~$\varphi(n)$ give rise to witnesses for~$\psi(n)$.

\begin{table}
  \centering
  \small
  \renewcommand{\arraystretch}{1.3}
  \begin{tabular}{@{}rp{4.3cm}p{5.3cm}@{}}
    \toprule
    & {classical logic} & {intuitionistic logic}
    \\\midrule
    statement $\varphi$ & The statement $\varphi$ holds. & We have a witness for $\varphi$. \\
    $\bot$ & A contradiction holds. & We have a witness for a contradiction. \\
    $\varphi \wedge \psi$ & $\varphi$ and $\psi$ hold. & We have a witness for~$\varphi$ and for~$\psi$. \\
    $\varphi \vee \psi$ & $\varphi$ or $\psi$ hold. & We have a witness for~$\varphi$ or for~$\psi$. \\
    $\varphi \Rightarrow \psi$ & If~$\varphi$ holds, then so does~$\psi$. &
    We can (uniformly) construct witnesses for~$\psi$ from witnesses
    for~$\varphi$. \\
    $\neg\varphi$ &
      $\varphi$ does not hold. &
      There is no witness for~$\varphi$. \\
    $\forall x\?X\_ \varphi(x)$ & For all $x \? X$ it holds that~$\varphi(x).$ &
      We can (uniformly) construct, for all~$x \? X$, witnesses for~$\varphi(x)$. \\
    $\exists x\?X\_ \varphi(x)$ & \raggedright There is at least one~$x \? X$
    such that~$\varphi(x)$ holds. & {\raggedright
      We have a~$x \? X$ together with a witness for~$\varphi(x)$.} \\
    \bottomrule
  \end{tabular}
  \caption{\label{table:bhk}Informal recursive definition of the notion of witnesses.}
\end{table}

\begin{ex}
According to the Brouwer--Heyting--Kolmogorov interpretation, the principle of
excluded middle states that we can construct, for every~$\varphi$,
either a witness for~$\varphi$ or a witness for~$\neg\varphi$. This claim is
obviously false.
\end{ex}

\begin{ex}
The BHK interpretation of a doubly negated statement~$\neg\neg\varphi$ is that
there is no witness for~$\neg\varphi$. This state of affairs does not entail
that we actually have a witness for~$\varphi$; in a sense, the
statement~$\varphi$ is only ``potentially true''.
\end{ex}

\begin{ex}
Assuming that we cannot find our apartment key, but know that it has to be
somewhere (because we used it last night to unlock the door), constructively we
can only defend the doubly negated statement
\[ \neg\neg (\exists x\_ \text{the key is at position~$x$}). \]
\end{ex}

\begin{ex}
We are running errands and are remembering that we need to fetch certain
ingredients. Unfortunately we don't recall any of them right now. Then we can
constructively defend the statement that the set of ingredients to fetch is not
empty, not however the stronger statement that this set is inhabited.
\end{ex}

\begin{ex}[\cite{sigfpe:katemoss,bbc:katemoss}]
A video surfaced depicting Kate Moss taking drugs, more precisely either drugs
of some type~A or drugs of some type~B. However from the video it wasn't clear
which type of drug it actually was. Hence there was no evidence for either type
of consumption; Kate Moss wasn't prosecuted.
\end{ex}

\begin{rem}The reference of an unspecified ``we'' in table~\ref{table:bhk}
renders our account of the BHK interpretation not only informal, but is also
somewhat misleading. As in classical mathematics, judgments of intuitionistic
logic do not actually depend on you, me or other mathematicians. It's perfectly
possible that statements which haven't yet been constructively verified admit
(as yet unknown) witnesses. More details can be found
in~\cite[pp.~42]{kohlenbach:applprooftheory} and
in~\cite{artemov:bhk,sanz-piecha:critical-bhk,dalen:bhk}. The
Brouwer--Heyting--Kolmogorov interpretation is made more formal and precise by
\emph{realizability theory}~\cite{bauer:realizability,oosten:realizability}.
\end{rem}

\subsection{The computability interpretation}

A second interpretation of mathematical statements suited for constructive
mathematics is given by the following motto: In a certain sense, \emph{we
accept a mathematical statement in constructive mathematics if and only if
there is a computer program witnessing it in finite time.} For instance, that
the statement
\[ \forall n \? \NN\_ \exists p \? \NN\_ p \geq n \wedge \text{$p$ is prime} \]
is valid in constructive mathematics corresponds to the fact that there is a
computer program which reads a number~$n$ as input and then, after some
computation, produces a prime number~$p \geq n$ as output (together with a
witness that~$p$ is in fact greater or equal than~$n$ and that~$p$ is prime).

For the motto to be correct, the notion of ``witnessed by a computer program''
has to be interpreted in sufficiently generous manner, not least because
uncountable and uncomputable structures have their place in constructive
mathematics just as in classical mathematics. A more formal rendition is
provided by the celebrated \emph{Curry--Howard correspondence} and
\emph{propositions as (some) types}.

\subsection{The geometric interpretation}

A third interpretation of mathematical statements suited for constructive
mathematics, and of a substantially different kind than the BHK and the
computability interpretation, is provided by geometry. In a certain sense,
\emph{we accept a mathematical statement in constructive mathematics if and
only if it holds locally in continuous families}.

For instance, while it is a fact of classical mathematics that every complex
number has a square root, a fundamental observation in complex analysis is that
such roots cannot be picked in a locally continuous manner. Hence the standard
formulation of the fundamental theorem of algebra fails in constructive
mathematics.\footnote{More precisely, the fundamental theorem of algebra fails
for the complex numbers built from pairs of Dedekind reals. Constructively, the
reals built using Dedekind cuts are not the same as the reals built using
Cauchy sequences; in fact, the latter are a subset of the former and serve a
different purpose. Such a bifurcation of classically equivalent notions is
typical of constructive mathematics. The fundamental theorem of algebra is
correct for the flavor of complex numbers built from Cauchy
reals~\cite{ruitenburg:roots}, and also for the algebraic
numbers~\cite[Section~VII.4]{mines-richman-ruitenburg:constructive-algebra}
[XXX insert better references] and for complex polynomials (of either flavor)
which are separable. Furthermore, the fundamental theorem of algebra holds in
an approximate sense and in a multiset sense~\cite{richman:fta}, and the
standard version only requires mild choice
principles~\cite{bridges-richman-schuster:wcc}.} This example will be discussed
in more detail in Section~\ref{sect:fta}.

The geometric interpretation not only gives transparent explanations to the
logical subtleties of constructive mathematics, but is in a more precise form
also the workhorse of many modern applications of constructive mathematics to
mathematics in general. We will hence dedicate ample of space to the geometric
interpretation by devoting Lecture~\ref{lect:toposes} to it, and will discuss
some of its more advanced applications in Lecture~\ref{lect:generic}.

\section{Finer distinctions supported by constructive mathematics}

@@

{\noindent\scriptsize\emph{Remark.} The set~$\NN_\infty$ is the \emph{one-point
compactification} of~$\NN$. A sensible definition of it in constructive mathematics
is as the set of decreasing binary sequences~$(x_0,x_1,x_2,\ldots)$. The
naturals embed into~$\NN_\infty$ by mapping~$n$ to the sequence~$1^n 0^\omega =
(1,\ldots,1,0,\ldots,0)$, and an element not in the image of this embedding
is~$\infty \defeq 1^\omega = (1,1,\ldots)$. Assuming the principle of excluded
middle (or already weaker principles), every element of~$\NN_\infty$ is of one
of these two forms. Martín Escardó has worked extensively on unexpected
instances of the principle of omniscience for searchable sets
like~$\NN_\infty$~\cite{escardo:omniscience1,escardo:omniscience2,escardo:omniscience}.\par}
\end{enumerate}
\end{exercise}

@@

\begin{exercise}[Brouwerian counterexamples]
Show that each of the following statements implies the principle of excluded middle,
hence is not available in constructive mathematics.
\begin{enumerate}
\item Every ideal of~$\ZZ$ is finitely generated.

@@

{\noindent\scriptsize\emph{Note.} A set~$X$ is \emph{Kuratowski-finite} if and
only if, for some number~$n \in \NN$, there is a surjective map~$[n] \to X$,
where~$[n] = \{ 0,1,\ldots,n-1 \}$. More briefly, a set~$X$ is
Kuratowski-finite iff its elements can be enumerated: $X = \{x_1,\ldots,x_n \}$.\par}
\end{enumerate}
Show that the following statement implies Markov's principle (from
Exercise~\ref{ex:tautologies}):
\begin{enumerate}
\addtocounter{enumi}{3}
\item Every real number which is not zero is invertible.
\end{enumerate}
\noindent
Brouwerian counterexamples abound in constructive mathematics; when developing
a constructive account of a theory, they help to clearly demarcate its limits.
As such additional Brouwerian counterexamples can be found in most texts on
constructive mathematics, such
as~\cite{mines-richman-ruitenburg:constructive-algebra}; a compilation mainly
from constructive analysis can be found
in~\cite{mandelkern:brouwerian-counterexamples}.
\end{exercise}

\begin{exercise}[Diaconescu's theorem]
\label{ex:diaconescu}
The axiom of choice can be put as: ``Every surjective map has a
section.'' (A \emph{section}~$s$ to a surjective map~$f$ is a map in the other direction such that~$f \circ s =
\mathrm{id}$.) A theorem of Diaconescu states that
the axiom of choice implies the principle of excluded middle. To this end, let~$\varphi$ be a
statement and consider the subsets
\begin{align*}
  U &= \{ x \in X \,|\, (x = 0) \vee \varphi \} \\
  V &= \{ x \in X \,|\, (x = 1) \vee \varphi \}
\end{align*}
of the discrete set~$X \defeq \{ 0,1 \}$.
\begin{enumerate}
\item Verify that~$U = V$ if and only if~$\varphi$.
\item Using that~$x = y \vee x \neq y$ for all elements~$x,y \in X$, show that
the existence of a section of the surjective map
\[ \begin{array}{@{}rcl@{}}
  X &\longrightarrow& \{U,V\} \\
  0 &\longmapsto& U \\
  1 &\longmapsto& V
\end{array} \]
implies~$\varphi \vee \neg\varphi$.
\end{enumerate}
\end{exercise}

% - What constructive mathematics is about: finer distinctions, algorithmic
%   interpretation, geometric interpretation (including teaser on why we
%   might care)
% - Proof by and proof of a contradiction: powers of irrational numbers,
%   discreteness of N, Q, Qbar, failure of discreteness of R
% - Positive information from negative? de Morgan, minimums of sets of
%   natural numbers
% - Relation to classical mathematics
% - Finite sets and the axiom of choice
% - Axiomatic freedom: all functions continuous, all functions
%   computable, generic prime ideal

\chapter{On the constructive content of classical proofs}
\label{lect:constructive-content}

\begin{intro}
Decades of experience in constructive mathematics show: \emph{Most results
in classical mathematics, even those whose proof rests on
non-constructive principles like the axiom of choice or the principle of
excluded middle, have a hidden constructive core.} With a mix of
experience, seasoned tools and general metatheorems, this constructive
content can be extracted from classical proofs. In this way we obtain
constructive reformulations of classical results, especially
if they are of a sufficiently concrete nature.

@@

What is the meaning of these non-effective proofs? Theorem~\ref{thm:dickson}
claims the existence of a finite object with a decidable property (a
pair~$(i,j)$ such that~$f(i) \leq f(j)$), but the given proof employs
transfinite methods and gives no indication how we could compute or otherwise
find this object. Instead, the classical proof asks us to grasp the infinitude
of all values of~$f$ so we can determine the minimum. The issue is even more
pronounced with Theorem~\ref{thm:surjective-matrix}, since the existence of
maximal ideals in nontrivial rings requires the axiom of
choice~\cite{scott:prime-ideals,hodges:krull,banaschewski:krull,erne:krull,howard-rubin:ac}.
To add one more conundrum: Assume that we have a classical proof, using the principle
of excluded middle and the axiom of choice, that some given Turing machine
terminates. Can we then constructively accept that the machine will halt? Do we
have an upper bound for the number of computational steps the machine carries
out before halting?

@@

\item \emph{Eliminating the principle of excluded middle by the double-negation
translation and its variants.} At the price of
slightly modifying the asserted claim, the principle of excluded middle
can always be mechanically eliminated from a given proof. This
elimination procedure is facilitated by the \emph{double-negation translation}
reviewed below. In some cases, a refined translation even allows to preserve the asserted
claim exactly; this technique is variously known as \emph{Friedman's trick},
\emph{nontrivial exit continuation} or (the baby version of) \emph{Barr's
theorem}. To cite a concrete instance of this phenomenon, classical~\textsc{zf}
set theory is conservative over its intuitionistic cousin~\textsc{izf}
for~$\Pi^0_2$-statements (statements of the form~$\forall\ldots\forall\_
\exists\ldots\exists\_ \%$, where all quantifiers in the final~``$\%$''
are bounded).

@@

Noticeably missing in this list is any technique for eliminating uses of the
\emph{powerset axiom} stating that the collection of all subsets of a given
set is again a set. While this axiom is uncontested by ordinary constructive
mathematics and doesn't receive nearly as much philosophical attention as the
axiom of choice or the principle of excluded middle, it is this axiom which actually
and substantially increases logical strength. While~\textsc{zfc}, \textsc{zf}
and \textsc{izf} are equiconsistent (and in fact verify the same
arithmetical~$\Pi^0_2$-statements), systems without it such as Kripke--Platek
set theory~(\textsc{kp}) or constructive Zermelo--Fraenkel set
theory~(\textsc{czf}) are much weaker. We will not discuss this curious state
of affairs; references include~\cite{crosilla:predicativity,aczel-rathjen:cst}.

\section{The double-negation translation}
\newpage

\section{Barr's theorem}
\newpage

@@

\section{Examples from quadratic form theory}
\newpage

@@

\chapter{Generic models and their applications}
\label{lect:generic}

@@

\section{A fantastical ring}
\newpage
\ \newpage

\section{The generic prime filter}
\newpage
\ \newpage

\section{Explicit construction}
\newpage
\ \newpage
